Interfacing logic programming and Coq

### 2020-10-29 papers/types2020/app.tex
added some related and future work, suggestions for title, revisions to sec 4

Stuff we may not have space for in the paper
\begin{itemize}
\item The LJF sequent rules
\item the code for
  \lsti{dep_pbt}

\item \dots
\end{itemize}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

In this appendix we list some definitions and pieces of code that may be helpful

\subsection{Interp}

In Fig.~\ref{fig:interp} we the encoding of the vanilla meta-interpreter used in
the testing phase of the \lsti{dep_pbt} tactic. Differently from Fig.~\ref{fig:augmented} we appeal, via quotations, to Coq's defined connectives --- note the use of a green cut to avoid overlapping with the \lsti{Inductive} atomic case and the absence of certificates.

\begin{figure}
% \newcommand{\XXi}{\Pscr}
% \newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
% \[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
% \infer{\XXi\vdash G_1\wedge G_2}
% {\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
% \]
% \vskip -18pt
% \[
% \infer{\XXi\vdash A}
%       {\bc D A \quad D \in \Pscr}\qquad
%   \infer{\bc A A}{}{}\qquad
%   \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
%   \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
% \]
% In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
% formulas.
 \lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{Encoding in coq-elpi of the vanilla meta-interpreter}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-10-29 papers/types2020/conc.tex
added some related and future work, suggestions for title, revisions to sec 4

A mention to the use of \emph{delays} in \texttt{check}: instead of generating only ground terms, we could generate say lists with at least two elements as \texttt{[1;2|Tail]} or simply not commit to a base type. This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or the stuff by Fabio Fioravanti).
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

Link to the repo
\subsection{Related work}
\label{ssec:rel}

\begin{itemize}
\item Brief discussion/comparison with \texttt{auto}
\item Mention other papers by Enrico as applications of coq-elpi
\item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in say MetaCoq would be a disaster (reimplement unfication, backtracking, binders)
\end{itemize}

\subsection{Future work}
\label{ssec:fut}

A mention to the use of \emph{delays} in \texttt{check}: instead of generating only ground terms, we could generate say lists with at least two elements as \texttt{[1;2|Tail]} or simply not commit to a base type. This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or the stuff by Fabio Fioravanti).

Extending our tactic to handle computations (\texttt{simp})

Extending the LJF kernel to handle defined atoms
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-10-29 papers/types2020/pbt.tex
added some related and future work, suggestions for title, revisions to sec 4

In a previous paper~\cite{blanco19ppdp}, we presented a
proof-theoretical reconstruction of property-based testing (PBT) of
relational specifications, adopting techniques from foundational proof
certificates to account for several features of these testing
paradigm: from various \emph{generation} strategies, to
\emph{shrinking} and fault localization.

Given the connection that \textsf{coq-elpi} offers between logic
programming and the internals of Coq, it is natural to extend the FPC-driven logic programming interpreter of the previous section to perform PBT of \lstinline{Inductive} types.

@@

Of course, \textsf{QuickChick}~\cite{QChick}
(\url{https://softwarefoundations.cis.upenn.edu/qc-current}) is a
sophisticated and far-reaching PBT tool for Coq, which is based on a
different perspective: being a clone of Haskell's QuickCheck, it
emphasizes testing executable (read \emph{decidable)} specifications
with random generators. While current research~\cite{LampropoulosPP18} aims to increase automation, it is fair to say testing, in particular of  relational specifications, is still very labor intensive. We do not intend to compete with  \textsf{QuickChick}, but as we shall see, as far as \lsti{Inductive} definitions that corresponds to pure Horn programs, we can test them immediately without any decidability proof or configuration of generators and shrinkers.

@@

\subsection{Implementing the tactic}
We will invoke the tactic in a
proof environment  where the goal is % one is trying to prove is the
property that the specification should meet. This means that, after
\lsti|intro| has been used to introduce the relevant hypotheses, we
will have a context % consisting of typing information for a bunch of variables
that we  divide in two categories: dependent
variables, which appear also in the types of other variables, and
non-dependent variables tat do not % Obviously, the
% two are the result of and \lsti|intro| on either a dependent or a
% non-dependent product
; in our setting this division acquires an additional reading beyond
what the property looks like: since specifications are encoded as pure
Horn clauses where no dependency is involved non-dependent variables
simply tag the specification clauses, whereas dependent variables
represent the typing information for the data variables that the
speficications use.

Whithin the tactic the user
specifies which variables of the environment should be used for
generating data and which for executing the specification. In addition
to this, the user should specify all the certificate information that
will guide the data generation phase. The goal of the current
environment will directly provide the property for which we want to
find a counterexample. % Say we want to generate a counterexample for
% some specification represented by \lsti|H1| and \lsti|H2| in the
% environment, and we want to generate the data for some variables
% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
% the tactic will then look like:
% \begin{lstlisting}
%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
% \end{lstlisting}
Formally, for a property
\[
\forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
\]
and the specification of a FPC, the call will be:
\begin{lstlisting}
elpi dep_pbt <fpc> $(A_1\land\dots\land A_m) \ (x_1) \dots (x_n)$.
\end{lstlisting}

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to proofs bounded
by at most $5$ unfoldings of the library \lsti{Inductive} definition
of \lsti{nat} and \lsti{list}. Since we did slip in an error (where?), our
tactic reports a counter-example, namely
\verb|(* Proof Term: [0, [0; 1; 0]] *)|

In order to generate the query to the \lsti|check| predicate, some
pre-processing is needed. In particular, we need to turn the data
variables into \lP logic variables when they appear inside a
specification, and to generate queries to \lsti|check| for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to realize this preprocessing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses.

>>>>

We have presented n a previous paper~\cite{blanco19ppdp} a
proof-theoretical reconstruction of property-based testing (PBT) of
relational specifications, adopting techniques from foundational proof
certificates to account for several features of this testing
paradigm: from various \emph{generation} strategies, to
\emph{shrinking} and fault localization.

Given the connection that \textsf{coq-elpi} offers between logic
programming and the internals of Coq, it is natural to extend the
FPC-driven logic programming interpreter of the previous section to
perform PBT over \lstinline{Inductive} types.

@@

Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
(\url{https://softwarefoundations.cis.upenn.edu/qc-current}) is a
sophisticated and well-supported PBT tool, based on a different
perspective: being a clone of Haskell's QuickCheck, it emphasizes
testing executable (read \emph{decidable)} specifications with random
generators. While current research~\cite{LampropoulosPP18} aims to
increase automation, it is fair to say that testing, in particular of
relational specifications, is still very labor intensive. We do not
intend to compete with \textsf{QuickChick}, but we shall see that we
can test them immediately \lsti{Inductive} definitions that
corresponds to pure Horn programs, without any decidability proof nor
commitment to the random generation strategy, and related
configuration of generators and shrinkers.

@@

\subsection{An Elpi tactic for PBT}
We will invoke the tactic in a
proof environment  where the goal is % one is trying to prove is the
the property that the system-under-test should meet. This means that, after
\lsti|intro| has been used to introduce the relevant hypotheses, we
 have a proof context % consisting of typing information for a bunch of variables
where eigenvariables  play different roles:  dependent
variables, that is the ones appearing also in the types of other variables, are subject to generation, while
non-dependent ones tags assumptions that are only to tested. This is reminiscent of bi.directional type-checking. % Obviously, the
% two are the result of and \lsti|intro| on either a dependent or a
% non-dependent product
% ; in our setting this division acquires an additional reading beyond
% what the property looks like: since specifications are encoded as pure
% Horn clauses where no dependency is involved, non-dependent variables
% simply tag the specification clauses, whereas dependent variables
% represent the typing information for the data variables that the
% specifications use.

Within the tactic the user therefore
specifies which variables of the environment should be used for
generating data and which for executing the specification. In addition
to this, the user should specify all the certificate information that
will guide the data generation phase. The goal of the current
environment will directly provide the property for which we want to
find a counterexample. % Say we want to generate a counterexample for
% some specification represented by \lsti|H1| and \lsti|H2| in the
% environment, and we want to generate the data for some variables
% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
% the tactic will then look like:
% \begin{lstlisting}
%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
% \end{lstlisting}
Concretely, for a property
\[
\forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
\]
and the specification of a FPC, the call to the tactic will be:
\begin{lstlisting}
elpi dep_pbt <fpc> $(A_1\land\dots\land A_m) \ (x_1) \dots (x_n)$.
\end{lstlisting}
The tactic will call \lsti{check} with the given FPC on the dependent variables and will simply call a vanilla meta-interpreter (see Fig.~\ref{fig:interp} in the appendix) to test
\begin{lstlisting}
interp $(A_1\land\dots\land A_m$), not (interp $B$)}
\end{lstlisting}
where \lsti{not} in \lP's negation-as-failure operator.

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is
whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to proofs bounded
by at most $5$ unfoldings of the library \lsti{Inductive} definition
of \lsti{nat} and \lsti{list}. Since we did slip in an error (where?), our
tactic reports a counter-example, namely
\verb|(* Proof Term: [0, [0; 1; 0]] *)|

In order to generate the query to the \lsti|check| predicate, some
pre-processing is needed. In particular, we need to turn the data
variables into \lP logic variables when they appear inside a
specification, and to generate queries to \lsti|check| for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to realize this pre-processing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses.

### 2020-10-29 papers/types2020/sec3.tex
added some related and future work, suggestions for title, revisions to sec 4

\newpage
\section{Proof theory and proof certificates}

@@

  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi;\Gamma\vdash t : A}
      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be produced.
}

@@

\begin{lstlisting}
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
}}.
\end{lstlisting}
The following example show how we can do FPC-driven logic programming in Coq and return
a Coq term
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog 10.
Qed.
Print i1.
\end{lstlisting}
\begin{verbatim}
ex_intro (fun R : list nat => insert 2 [0; 1] R) [2; 2; 2]
  (i_c 2 0 [1] [2; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
     : exists R : list nat, insert 2 [0; 1] R
\end{verbatim}
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

%\newpage
\section{Proof theory and proof certificates}

@@

  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi;\Gamma\vdash t : A}
      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{blanco17cade}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be produced.
}

@@

\begin{lstlisting}
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
}}.
\end{lstlisting}
The following example show how we can do FPC-driven logic programming in Coq and return
a Coq term
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog 10.
Qed.
Print i1.
\end{lstlisting}
\begin{verbatim}
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)))
	 : exists R : list nat, insert 2 [0; 1] R
\end{verbatim}
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
